POPLMark goes Concurrent!\\Challenge problems for process calculi and behavioural types

### 2023-10-11 challenges/linearity.tex
pass on linearity appendix

\subsubsection{Syntax}
The syntax is given by the grammar:
\[
\begin{array}{rcl}
  v,w & ::= & a \mid l \\
   P,Q & ::=& \Pend \mid \Pout{x}{v}{P} \mid \Pin{x}{l}{P} \mid \PBpar{P}{Q} \mid  \Presd{x}{y}{P}
\end{array}
\]
A \emph{value} \( v, w, \dots \) is either a base value \( a \) or a variable name \( l \).

The output process \( \Pout{x}{v}{P} \) can send the value \( v \) via \( x \), then continue as \( P \).
The intention is that the value \( v \) must be a base value when it is actually sent, and this will be enforced in the semantics later on.
The input process \( \Pin{x}{l}{P} \) is an \emph{input}, which can receive a base value via \( x \), then continue as \( P \) with the received value substituted for the variable name \( l \).

The process \( \Presd{x}{y}{P} \) is the \emph{restriction} of the channel with endpoints named \( x \) and \( y \) to \( P \).
Components in \( P \) can use the names \( x \) and \( y \) to interact with each other (sending on \( x \) and receiving on \( y \) or vice versa), but not with processes outside of the restriction.
The restriction operator thus binds the names \( x \) and \( y \) in \( P \).
Note that the scope of a restriction may not change when processes interact, since it is only possible to send and receive values, and not names.
Note that there is no recursion of replication in the syntax, and thus no infinite behaviors can be expressed.

\subsubsection{Semantics}
We describe the actions that the system can perform using a small step operational semantics.
To simplify the definition of the reduction relation, we will  factor out a \emph{structural congruence} relation on processes, namely  the smallest congruence relation that satisfies the following axioms:
\begin{mathpar}
  \inferrule[Sc-Par-Comm]{ }{\scong{\Ppar{P}{Q}}{\Ppar{Q}{P}}}
  \and
  \inferrule[Sc-Par-Assoc]{ }{\scong{\Ppar{(\Ppar{P}{Q})}{R}}{\Ppar{P}{(\Ppar{Q}{R})}}}
  \and
  \inferrule[Sc-Par-Inact]{ }{\scong{\Ppar{P}{\Pend}}{P}}
  \and
  \inferrule[Sc-Res-Par]{\{x,y\} \cap \freenames{Q} = \emptyset}{\scong{\Ppar{\Presd{x}{y}{P}}{Q}}{\Presd{x}{y}{(\Ppar{P}{Q})}}}
  \and
  \inferrule[Sc-Res-Inact]{ }{\scong{\Presd{x}{y}{\Pend}}{\Pend}}
  \and
  \inferrule[Sc-Res]{ }{\scong{\Presd{x_1}{y_1}{\Presd{x_2}{y_2}{P}}}{\Presd{x_2}{y_2}{\Presd{x_1}{y_1}{P}}}}
\end{mathpar}

The operational semantics is then defined as the following binary relation on processes:
\begin{mathpar}
  \inferrule[R-Com]{ }{\reduces{\Presd{x}{y}{(\Ppar{\Pout{x}{a}{P}}{\Ppar{\Pin{y}{l}{Q}}{R}})}}{\Presd{x}{y}{(\Ppar{P}{\Ppar{\subst{Q}{a}{l}}{R}})}}}
  \and
  \inferrule[R-Res]{\reduces{P}{Q}}{\reduces{\Presd{x}{y}{P}}{\Presd{x}{y}{Q}}}
  \and
  \inferrule[R-Par]{\reduces{P}{Q}}{\reduces{\Ppar{P}{R}}{\Ppar{Q}{R}}}
  \and
  \inferrule[R-Struct]{\scong{P}{P'} \\ \reduces{P'}{Q'} \\ \scong{Q}{Q'}}{\reduces{P}{Q}}
\end{mathpar}
Note that there is no rule for inferring an action of an input or output process except those that match the input/output capability.
Note also that due to rule \TirName{R-Com}, the process \( \Pin{y}{l}{P} \) can receive \emph{any} base value.
On the other hand, since the rule \TirName{R-Com} only applies to sending base values, there is no way to send a variable name or a name.

\subsubsection{Session types}
Our process syntax allows us to write processes that are not well-formed in the sense that they either use the endpoints bound by a restriction to communicate in a way that does not follow the intended duality of communication, or attempt to send something which is not a base value.
As an example, the following process attempts to send a base value on both \( x \) and \( y\), whereas the intention of restricting names is that one of the names is used for sending and the other for receiving: \( \Presd{x}{y}{(\Ppar{\Pout{x}{a}{\Pend}}{\Pout{y}{a}{\Pend}})} \).
Another example is the following process, which attempts to send a variable that is not instantiated at the time of sending: \( \Presd{x}{y}{(\Ppar{\Pout{x}{l}{\Pend}}{\Pin{y}{l}{\Pend}})} \).

To alleviate this issue, we introduce a \emph{session type system}, which will detect processes that are not well-formed.

\paragraph{Syntax}
Our type system does not type processes directly, but instead focuses on the channels used in the process.
The syntax of \emph{session types} \( S, T \), unrestricted type contexts \( \Gamma \) and linear type contexts \( \Delta \) is as follows:
\[
  \begin{array}{rcl}
  S,T & ::= & \Tend \mid \Tbase \mid \Tin{S} \mid \Tout{S} \\
    \Gamma &::= & \Cempty \mid \Gamma, l \\
                    \Delta &::= & \Cempty \mid \Cadd{\Delta}{\hastype{x}{S}}
  \end{array}
\]
The \emph{end type}, \( \Tend \), is the type of an endpoint on which no further interaction is possible.
The \emph{base type}, \( \Tbase \), is the type of base values.
The \emph{input type}, \( \Tin{S} \), is the type of an endpoint which is ready to receive a value, then continue with type \( S \).
The \emph{output type}, \( \Tout{S} \), is the type of an endpoint which is ready to send a value, then continue with type \( S \).

Typing contexts gather type information about names and variables.
We distinguish between an unrestricted context $\Gamma$ seen as a set: since  we have only one base type, we only record which variable names have been used.
Type information for endpoints is captured in the \emph{linear}
context \( \Delta \). We use the comma as split/union, overloaded to
singletons, and \( \Cempty \) is the empty multiset.
We extend the Barendregt convention to contexts, so that all entries
are distinct.  Note that the order in which information is added to a
type context thus does not matter.

Since we need to determine whether endpoints are used dually to determine whether processes are well-formed, we will need to formally define the dual of a type.
The dual of a type, \( \dual{T} \), is defined by the following equations:
\begin{mathpar}
  \inferrule{}{\dual{\Tin{S}} = \Tout{\dual{S}}}
  \and
  \inferrule{}{\dual{\Tout{S}} = \Tin{\dual{S}}}
  \and
  \inferrule{}{\dual{\Tend} = \Tend}
\end{mathpar}
Note that the dual function is not total, since it is not defined for the base type.

\subsubsection{Typing rules}
Our type system will maintain two invariants:
\begin{enumerate}
\item Endpoints are used exactly once
\item Endpoints that are part of the same restriction have dual types
\end{enumerate}
The first invariant is maintained by linearly splitting type contexts when typing compositions of processes.
The second invariant is maintained by requiring duality when typing restrictions.

@@

\subsubsection{Challenge}
The objective of this challenge is to prove subject reduction and type safety for our calculus with session types. We start with:
\begin{theorem}[Subject reduction]
  If \( \types{\Gamma;\Delta}{P} \) and \( \reduces{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{theorem}
\begin{proof}
  Induction on the reduction relation, TBD.
\end{proof}

@@

To formulate safety, we need to formally define what we mean by well-formed.
We say that a process \( P \) is \emph{prefixed at variable \( x \)} if it is of the form \( \Pout{x}{v}{P} \) or \( \Pin{x}{l}{P} \).
A process is then \emph{well-formed} if, for each of its structurally congruent processes of the form \( \Presd{x_1}{y_1}{\dots \Presd{x_n}{y_n}{(\Ppar{\Ppar{P}{Q}}{R})}} \), with \( n \geq 0 \), it holds that, if \( P \) is prefixed at \( x_1 \) and \( Q \) is prefixed at \( y_1 \), then \( \Ppar{P}{Q} \) is of the form \( \Ppar{\Pout{x_1}{a}{P'}}{\Pin{y_1}{l}{Q'}} \).

Note that well-formed processes do not necessarily reduce, since e.g.\ the process
\begin{equation*}
  \Presd{x_1}{y_1}{\Presd{x_2}{y_2}{(\Ppar{\Pout{x_1}{a}{\Pin{y_2}{l}{\Pend}}}{\Pout{y_2}{x_2}{\Pin{y_1}{l}{\Pend}}})}}
\end{equation*}
is well-formed, but has no reduction.

\begin{theorem}[Type safety]
  If \( \types{\Cempty}{P} \), then \( P \) is well-formed.
\end{theorem}

\begin{corollary}
  If \( \types{\Cempty}{P} \) and \( \reduces{P}{Q} \) then \( Q \) is well-formed.
\end{corollary}

>>>>

\subsubsection{Syntax.}
The syntax is given by the grammar
\[
\begin{array}{rcl}
  v,w & ::= & a \mid l \\
   P,Q & ::=& \Pend \mid \Pout{x}{v}{P} \mid \Pin{x}{l}{P} \mid \PBpar{P}{Q} \mid  \Presd{x}{y}{P}
\end{array}
\]
where a \emph{value} \( v, w, \dots \) is either a base value \( a \) or a variable name \( l \).

The output process \( \Pout{x}{v}{P} \) sends the value \( v \) via \( x \) and then continues as \( P \).
The intention is that the value \( v \) must be a base value when it is actually sent, and this will be enforced in the semantics later on.
The input process \( \Pin{x}{l}{P} \) waits for a base value from \( x \) and then continues as \( P \) with the received value substituted for the variable name \( l \).
%
The process \( \Presd{x}{y}{P} \) represents a \emph{session} with endpoints named \( x \) and \( y \) which are bound in \( P \). In \( P \), the names \( x \) and \( y \) can be used to exchange messages over the session (sending on \( x \) and receiving on \( y \) or vice versa).
%
Finally, the process $\PBpar{P}{Q}$ represents the parallel composition of $P$ and $Q$.
%
Note that in this calculus channels cannot be sent in messages, therefore the topology of the communication network described by a process cannot change.
Also, there is no recursion or replication in the syntax, hence no infinite behaviors can be expressed. In particular, we only model linear (as opposed to shared) channels.

\subsubsection{Semantics.}
We describe the actions that the system can perform through a small step operational semantics.
As usual, we make use of a \emph{structural congruence} relation that equates processes that we deem to be indistinguishable. Structural congruence is the smallest congruence relation that satisfies the following axioms:
\begin{mathpar}
  \inferrule[Sc-Par-Comm]{ }{\scong{\Ppar{P}{Q}}{\Ppar{Q}{P}}}
  \and
  \inferrule[Sc-Par-Assoc]{ }{\scong{\Ppar{(\Ppar{P}{Q})}{R}}{\Ppar{P}{(\Ppar{Q}{R})}}}
  \and
  \inferrule[Sc-Par-Inact]{ }{\scong{\Ppar{P}{\Pend}}{P}}
  \and
  \inferrule[Sc-Res-Par]{\{x,y\} \cap \freenames{Q} = \emptyset}{\scong{\Ppar{\Presd{x}{y}{P}}{Q}}{\Presd{x}{y}{(\Ppar{P}{Q})}}}
  \and
  \inferrule[Sc-Res-Inact]{ }{\scong{\Presd{x}{y}{\Pend}}{\Pend}}
  \and
  \inferrule[Sc-Res]{ }{\scong{\Presd{x_1}{y_1}{\Presd{x_2}{y_2}{P}}}{\Presd{x_2}{y_2}{\Presd{x_1}{y_1}{P}}}}
\end{mathpar}

The operational semantics is then defined as the following binary relation on processes:
\begin{mathpar}
  \inferrule[R-Com]{ }{\reduces{\Presd{x}{y}{(\Ppar{\Pout{x}{a}{P}}{\Ppar{\Pin{y}{l}{Q}}{R}})}}{\Presd{x}{y}{(\Ppar{P}{\Ppar{\subst{Q}{a}{l}}{R}})}}}
  \and
  \inferrule[R-Res]{\reduces{P}{Q}}{\reduces{\Presd{x}{y}{P}}{\Presd{x}{y}{Q}}}
  \and
  \inferrule[R-Par]{\reduces{P}{Q}}{\reduces{\Ppar{P}{R}}{\Ppar{Q}{R}}}
  \and
  \inferrule[R-Struct]{\scong{P}{P'} \\ \reduces{P'}{Q'} \\ \scong{Q}{Q'}}{\reduces{P}{Q}}
\end{mathpar}

Note that there is no rule for inferring an action of an input or output process except those that match the input/output capability.\luca{Unclear to me. Since it is a reduction semantics it's obvious that input/output actions alone cannot do anything. I would remove this sentence}
%
Note that reductions are allowed only for restricted pairs of session endpoints. This makes it possible to formulate subject reduction so that the typing context is exactly the same before and after each reduction.
%
Note also that due to rule \TirName{R-Com}, the process \( \Pin{y}{l}{P} \) can receive \emph{any} base value.
On the other hand, since the rule \TirName{R-Com} only applies to sending base values, there is no way to send a variable name or a name.

\subsubsection{Session types.}
Our process syntax allows us to write processes that are ill formed in the sense that they either use the endpoints bound by a restriction to communicate in a way that does not follow the intended duality of communication, or attempt to send something which is not a base value.
As an example, the process \( \Presd{x}{y}{(\Ppar{\Pout{x}{a}{\Pend}}{\Pout{y}{a}{\Pend}})} \) attempts to send a base value on both \( x \) and \( y\), whereas one of the names should be used for receiving in order to guarantee progress.
Another example is the process \( \Presd{x}{y}{(\Ppar{\Pout{x}{l}{\Pend}}{\Pin{y}{l}{\Pend}})} \), which attempts to send a variable that is not instantiated at the time of sending.
%
To prevent these issues, we introduce a \emph{session type system} which
rules out ill-formed processes.

\paragraph{Syntax}
Our type system does not type processes directly, but instead focuses on the channels used in the process.
The syntax of \emph{session types} \( S, T \), unrestricted type contexts \( \Gamma \) and linear type contexts \( \Delta \) is as follows:
\[
  \begin{array}{rcl}
  S,T & ::= & \Tend \mid \Tbase \mid \Tin{S} \mid \Tout{S} \\
    \Gamma &::= & \Cempty \mid \Gamma, l \\
                    \Delta &::= & \Cempty \mid \Cadd{\Delta}{\hastype{x}{S}}
  \end{array}
\]
The \emph{end type} \( \Tend \) describes an endpoint on which no further interaction is possible.
The \emph{base type} \( \Tbase \) describes base values.
The \emph{input type} \( \Tin{S} \) describes endpoints used for receiving a value and then according to \( S \).
The \emph{output type} \( \Tout{S} \) describes endpoints used for sending a value and then according to \( S \).

Typing contexts gather type information about names and variables.
\emph{Unrestricted} contexts are simply sets of names since we only have one
base type. \emph{Linear contexts} associate a type to endpoints. We use
the comma as split/union, overloaded to singletons, and \( \Cempty \) is the
empty multiset. We extend the Barendregt convention to contexts, so that all
entries are distinct.  Note that the order in which information is added to a
type context does not matter.

Since we need to determine whether endpoints are used in complementary ways to determine whether processes are well formed, we need to formally define the dual of a type as follows:
\begin{mathpar}
  \inferrule{}{\dual{\Tin{S}} = \Tout{\dual{S}}}
  \and
  \inferrule{}{\dual{\Tout{S}} = \Tin{\dual{S}}}
  \and
  \inferrule{}{\dual{\Tend} = \Tend}
\end{mathpar}
Note that the dual function is partial since it is undefined for the base type.

\subsubsection{Typing rules}
Our type system is aimed at maintaining two invariants:
\begin{enumerate}
\item No endpoint is used simultaneously by parallel processes;
\item The two endpoints of the same session have dual types.
\end{enumerate}
The first invariant is maintained by linearly splitting type contexts when typing compositions of processes.
The second invariant is maintained by requiring duality when typing restrictions.

@@

\subsubsection{Challenge}
The objective of this challenge is to prove subject reduction and type safety for our calculus with session types. We start with:
\begin{theorem}[Subject reduction]
  If \( \types{\Gamma;\Delta}{P} \) and \( \reduces{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reduces{P}{Q}$. TBD
\end{proof}

@@

To formulate safety, we need to formally define what we mean by well-formed process.
We say that a process \( P \) is \emph{prefixed at variable \( x \)} if it is of the form \( \Pout{x}{v}{P} \) or \( \Pin{x}{l}{P} \).
A process is then \emph{well formed} if, for each of its structurally congruent processes of the form \( \Presd{x_1}{y_1}{\dots \Presd{x_n}{y_n}{(\Ppar{\Ppar{P}{Q}}{R})}} \), with \( n \geq 0 \), it holds that, if \( P \) is prefixed at \( x_1 \) and \( Q \) is prefixed at \( y_1 \), then \( \Ppar{P}{Q} \) is of the form \( \Ppar{\Pout{x_1}{a}{P'}}{\Pin{y_1}{l}{Q'}} \).

Note that well-formed processes do not necessarily reduce. For example, the process
\begin{equation*}
  \Presd{x_1}{y_1}{\Presd{x_2}{y_2}{(\Ppar{\Pout{x_1}{a}{\Pin{y_2}{l}{\Pend}}}{\Pout{y_2}{x_2}{\Pin{y_1}{l}{\Pend}}})}}
\end{equation*}
is well formed but also irreducible.

\begin{theorem}[Type safety]
  If \( \types{\Cempty}{P} \), then \( P \) is well formed.
\end{theorem}

\begin{corollary}
  If \( \types{\Cempty}{P} \) and \( \reduces{P}{Q} \) then \( Q \) is well formed.
\end{corollary}
